Project Report\\\textsc{DEVS Ex Machina}

### 2015-06-14 manual/projectrapport.tex
Update project report (benchmarks)

\chapter{Performance analysis}
To measure the performance of \textsc{DEVS Ex Machina} under more demanding conditions, \textbf{DEVStone} and \textbf{PHOLD} benchmarks were used in combination with the \texttt{perf} profiler tool. Every time we compare the performance of our implementation with \textit{PythonPDEVS}.\\
Note: at the moment, no major research has gone into the use of compiler optimizations yet. Potentially, there is still speed to be gained here.\\
\\
{\footnotesize[All the following tests were performed using an \textit{Intel Core i5 M520} CPU at 2.40GHz, parallel testing was done using 2 simulation cores.]}

\section{Classic DEVS}
Using Classic DEVS we see noticeable speedups compared to PyPDEVS. Both tests are completed in far less time, using the CPU more efficiently.

\subsection{DEVStone}
The DEVStone benchmark used for both implementations has a width of 2 and a depth of 3.
\paragraph{\textsc{DEVS Ex Machina}}
\begin{Verbatim}[fontsize=\small]
$ perf stat ./dxexmachina_devstone classic 2 3

 Performance counter stats for './dxexmachina_devstone classic 2 3':

        174.007707 task-clock (msec)         #    1.111 CPUs utilized          
               184 context-switches          #    0.001 M/sec                  
               212 cpu-migrations            #    0.001 M/sec                  
               696 page-faults               #    0.004 M/sec                  
       463,684,386 cycles                    #    2.665 GHz                     [83.13%]
       196,709,921 stalled-cycles-frontend   #   42.42% frontend cycles idle    [85.77%]
       103,770,146 stalled-cycles-backend    #   22.38% backend  cycles idle    [68.62%]
       557,355,313 instructions              #    1.20  insns per cycle        
                                             #    0.35  stalled cycles per insn [86.11%]
       131,411,927 branches                  #  755.208 M/sec                   [85.62%]
         1,720,031 branch-misses             #    1.31% of all branches         [83.38%]

       0.156664362 seconds time elapsed
\end{Verbatim}
\paragraph{\textsc{PythonPDEVS}}
\begin{Verbatim}[fontsize=\small]
$ perf stat python experiment.py default 2 false > out.txt

 Performance counter stats for 'python experiment.py default 2 false':

        709.776558 task-clock (msec)         #    0.989 CPUs utilized          
               191 context-switches          #    0.269 K/sec                  
                34 cpu-migrations            #    0.048 K/sec                  
             2,942 page-faults               #    0.004 M/sec                  
     1,687,298,167 cycles                    #    2.377 GHz                     [82.71%]
       866,759,527 stalled-cycles-frontend   #   51.37% frontend cycles idle    [83.25%]
       562,247,464 stalled-cycles-backend    #   33.32% backend  cycles idle    [67.62%]
     1,672,919,036 instructions              #    0.99  insns per cycle        
                                             #    0.52  stalled cycles per insn [83.82%]
       375,443,818 branches                  #  528.961 M/sec                   [83.25%]
        14,867,814 branch-misses             #    3.96% of all branches         [83.28%]

       0.717576473 seconds time elapsed
\end{Verbatim}
\subsection{PHOLD}
\paragraph{\textsc{DEVS Ex Machina}}
\begin{Verbatim}[fontsize=\small]
$ perf stat ./dxexmachina_phold classic 1 10 0 10

 Performance counter stats for './dxexmachina_phold classic 1 10 0 10':

         15.847971 task-clock (msec)         #    1.047 CPUs utilized          
                15 context-switches          #    0.946 K/sec                  
                28 cpu-migrations            #    0.002 M/sec                  
               710 page-faults               #    0.045 M/sec                  
        27,365,192 cycles                    #    1.727 GHz                     [74.72%]
        12,605,461 stalled-cycles-frontend   #   46.06% frontend cycles idle    [73.78%]
         7,424,569 stalled-cycles-backend    #   27.13% backend  cycles idle    [86.04%]
        31,095,649 instructions              #    1.14  insns per cycle        
                                             #    0.41  stalled cycles per insn
         6,955,945 branches                  #  438.917 M/sec                  
           132,996 branch-misses             #    1.91% of all branches         [74.99%]

       0.015131930 seconds time elapsed
\end{Verbatim}
\paragraph{\textsc{PythonPDEVS}}
\begin{Verbatim}[fontsize=\small]
$ perf stat python experiment.py default 2 false > out.txt

 Performance counter stats for 'python experiment.py default 2 false':

        709.776558 task-clock (msec)         #    0.989 CPUs utilized          
               191 context-switches          #    0.269 K/sec                  
                34 cpu-migrations            #    0.048 K/sec                  
             2,942 page-faults               #    0.004 M/sec                  
     1,687,298,167 cycles                    #    2.377 GHz                     [82.71%]
       866,759,527 stalled-cycles-frontend   #   51.37% frontend cycles idle    [83.25%]
       562,247,464 stalled-cycles-backend    #   33.32% backend  cycles idle    [67.62%]
     1,672,919,036 instructions              #    0.99  insns per cycle        
                                             #    0.52  stalled cycles per insn [83.82%]
       375,443,818 branches                  #  528.961 M/sec                   [83.25%]
        14,867,814 branch-misses             #    3.96% of all branches         [83.28%]

       0.717576473 seconds time elapsed
\end{Verbatim}

\section{Parallel DEVS}
Also in parallel simulations \textsc{DEVS Ex Machina} completes quite a bit quicker, again achieving higher CPU utilization.
Additionally we can see that Conservative PDEVS performs consistently significantly faster than Parallel PDEVS, in the case of PHOLD.

\subsection{DEVStone}
\paragraph{\textsc{DEVS Ex Machina} - Optimistic}
\begin{Verbatim}[fontsize=\small]
$ perf stat ./dxexmachina_devstone opdevs 2 2 3

 Performance counter stats for './dxexmachina_devstone opdevs 2 2 3':

         95.265846 task-clock (msec)         #    0.499 CPUs utilized          
               191 context-switches          #    0.002 M/sec                  
                15 cpu-migrations            #    0.157 K/sec                  
             1,229 page-faults               #    0.013 M/sec                  
       197,702,110 cycles                    #    2.075 GHz                     [82.53%]
        95,414,988 stalled-cycles-frontend   #   48.26% frontend cycles idle    [83.54%]
        47,283,951 stalled-cycles-backend    #   23.92% backend  cycles idle    [76.71%]
       199,327,993 instructions              #    1.01  insns per cycle        
                                             #    0.48  stalled cycles per insn [91.61%]
        47,887,986 branches                  #  502.677 M/sec                   [86.14%]
           931,688 branch-misses             #    1.95% of all branches         [75.36%]

       0.191066656 seconds time elapsed
\end{Verbatim}
\paragraph{\textsc{DEVS Ex Machina} - Conservative}
\begin{Verbatim}[fontsize=\small]
$ perf stat ./dxexmachina_devstone cpdevs 2 2 3

 Performance counter stats for './dxexmachina_devstone cpdevs 2 2 3':

        129.292299 task-clock (msec)         #    0.555 CPUs utilized          
               672 context-switches          #    0.005 M/sec                  
                15 cpu-migrations            #    0.116 K/sec                  
             2,177 page-faults               #    0.017 M/sec                  
       308,983,665 cycles                    #    2.390 GHz                     [74.75%]
       151,616,998 stalled-cycles-frontend   #   49.07% frontend cycles idle    [84.43%]
        82,891,133 stalled-cycles-backend    #   26.83% backend  cycles idle    [76.81%]
       276,296,034 instructions              #    0.89  insns per cycle        
                                             #    0.55  stalled cycles per insn [88.68%]
        66,513,459 branches                  #  514.443 M/sec                   [87.86%]
         1,390,075 branch-misses             #    2.09% of all branches         [77.49%]

       0.233046274 seconds time elapsed
\end{Verbatim}
\paragraph{\textsc{PythonPDEVS}}
\begin{Verbatim}[fontsize=\small]
$ perf stat mpirun -np 2 python experiment.py default 2 false > out.txt

 Performance counter stats for 'mpirun -np 2 python experiment.py default 2 false':

        787.125992 task-clock (msec)         #    0.371 CPUs utilized          
             1,120 context-switches          #    0.001 M/sec                  
                91 cpu-migrations            #    0.116 K/sec                  
             8,684 page-faults               #    0.011 M/sec                  
     1,786,217,349 cycles                    #    2.269 GHz                     [83.85%]
       935,480,076 stalled-cycles-frontend   #   52.37% frontend cycles idle    [84.18%]
       619,366,137 stalled-cycles-backend    #   34.67% backend  cycles idle    [65.26%]
     1,796,463,414 instructions              #    1.01  insns per cycle        
                                             #    0.52  stalled cycles per insn [84.16%]
       403,042,819 branches                  #  512.044 M/sec                   [84.70%]
        13,210,328 branch-misses             #    3.28% of all branches         [82.68%]

       2.123888919 seconds time elapsed
\end{Verbatim}

\subsection{PHOLD}
\paragraph{\textsc{DEVS Ex Machina} - Optimistic}
\begin{Verbatim}[fontsize=\small]
$ perf stat ./dxexmachina_phold opdevs 2 1 10 0 10

 Performance counter stats for './dxexmachina_phold opdevs 2 1 10 0 10':

         20.741980 task-clock (msec)         #    0.092 CPUs utilized          
                73 context-switches          #    0.004 M/sec                  
                 8 cpu-migrations            #    0.386 K/sec                  
               760 page-faults               #    0.037 M/sec                  
        28,281,718 cycles                    #    1.364 GHz                     [42.03%]
        14,899,039 stalled-cycles-frontend   #   52.68% frontend cycles idle    [79.62%]
         8,373,668 stalled-cycles-backend    #   29.61% backend  cycles idle   
        29,601,476 instructions              #    1.05  insns per cycle        
                                             #    0.50  stalled cycles per insn
         6,754,172 branches                  #  325.628 M/sec                  
           135,220 branch-misses             #    2.00% of all branches         [81.22%]

       0.225834153 seconds time elapsed
\end{Verbatim}
\paragraph{\textsc{DEVS Ex Machina} - Conservative}
\begin{Verbatim}[fontsize=\small]
$ perf stat ./dxexmachina_phold cpdevs 2 1 10 0 10

 Performance counter stats for './dxexmachina_phold cpdevs 2 1 10 0 10':

         28.509489 task-clock (msec)         #    0.286 CPUs utilized          
                95 context-switches          #    0.003 M/sec                  
                11 cpu-migrations            #    0.386 K/sec                  
               813 page-faults               #    0.029 M/sec                  
        54,757,291 cycles                    #    1.921 GHz                     [63.45%]
        29,057,227 stalled-cycles-frontend   #   53.07% frontend cycles idle    [86.87%]
        15,602,463 stalled-cycles-backend    #   28.49% backend  cycles idle    [90.08%]
        53,326,834 instructions              #    0.97  insns per cycle        
                                             #    0.54  stalled cycles per insn
        12,449,801 branches                  #  436.690 M/sec                  
           219,155 branch-misses             #    1.76% of all branches         [64.68%]

       0.099591165 seconds time elapsed
\end{Verbatim}
\paragraph{\textsc{PythonPDEVS}}
PHOLD in parallel on PyPDEVS crashed immediately with an MPI error, hence a comparison was not possible for now.

>>>>

\chapter{Performance analysis}
To measure the performance of \textsc{DEVS Ex Machina} under more demanding conditions, \textbf{DEVStone} and \textbf{PHOLD} benchmarks were used in combination with the \texttt{perf} profiler tool. As these are standardized tests, they allow easy comparison between our implementation and \textit{PythonPDEVS}.\\
Note: at the moment, no major research has gone into the use of compiler optimizations yet. Potentially, there is still speed to be gained here.\\
\\
{\footnotesize[All the following tests were performed using an \textit{Intel Core i5 M520} CPU at 2.40GHz, parallel testing was done using 2 simulation cores. All tests were ran 5 times to give an indication of the predictability of the result.]}

\pagebreak

\section{DEVStone}
\begin{figure}
\centering
\begin{minipage}{.5\textwidth}
  \centering
  \hspace*{-1.5in}
  \includegraphics[width=1.6\linewidth]{devstone_classic.png}
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
  \includegraphics[width=1.6\linewidth]{devstone_parallel.png}
\end{minipage}
\end{figure}

In both Classic and Parallel DEVS there are noticeable speedups compared to \textit{PyPDEVS}.\\
Looking at \textsc{DEVS Ex Machina} specifically we can see that in the first few, rather lightweight cases the single core Classic DEVS simulation is faster than PDEVS, in the first case even twice as fast. Once more models start to get involved however, things quickly turn around since the benefit of having the raw processing power of more cores available outweighs the cost of setting up and managing multiple threads.\\
Additionally, the final and most demanding test case, a DEVStone benchmark of depth and width 12, was also executed with 4 cores, which showed a slight improvement over using 2 cores.
\\

However, it should be noted that in PDEVS a somewhat thought-out assignment of models to cores is of critical importance. A parallel simulation with models placed haphazardly can perform \emph{significantly} worse than a Classic DEVS which doesn't have to deal with placement. In general, it is best to keep the most used connections intra-core.

\pagebreak

\section{PHOLD}
\begin{figure}
\centering
\begin{minipage}{.5\textwidth}
  \centering
  \hspace*{-1.5in}
  \includegraphics[width=1.6\linewidth]{phold_classic.png}
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
  \includegraphics[width=1.6\linewidth]{phold_parallel.png}
\end{minipage}
\end{figure}

{\footnotesize[ To make a useful comparison with \textit{Python PDEVS} possible, care was taken to make sure the "meaningless" computations related to the payload size were not removed by compiler optimisations ]}\\
\\
This benchmark tests the effect of different payload sizes in a highly interconnected environment.\\
In both environments \textsc{DEVS Ex Machina} is significantly faster than \textit{PyPDEVS}.
Especially interesting is the second graph with parellel results. Whereas in DEVStone Optimistic and Conservative DEVS performed relatively equally, here we can spot a clear discrepancy between them.\\

In Optimistic we can see that as the payload grows, the total runtime becomes less predictable. Because of the inherent interconnectivity of the benchmark, and because processing a message with a larger payload can take a while, the chance of reverts or even series of reverts is increased.

In Conservative simulation however, the flow time is tightly controlled, avoiding reverts. In this case this behavior leads to significantly better results.

\pagebreak

\section{Detailed stats}
Some more detailed statistics provided by \texttt{perf} are shown here. Each time, only the result of a single run is provided, to give a general indication.

\subsection{DEVStone}
\paragraph{\textsc{DEVS Ex Machina}}
\begin{Verbatim}[fontsize=\small]
$ perf stat ./dxexmachina_devstone classic 2 3
 Performance counter stats for './dxexmachina_devstone classic 2 3':

        189.176880 task-clock (msec)         #    1.115 CPUs utilized          
               243 context-switches          #    0.001 M/sec                  
               212 cpu-migrations            #    0.001 M/sec                  
               691 page-faults               #    0.004 M/sec                  
       476,733,288 cycles                    #    2.520 GHz                     [84.35%]
       204,107,284 stalled-cycles-frontend   #   42.81% frontend cycles idle    [84.40%]
        95,948,277 stalled-cycles-backend    #   20.13% backend  cycles idle    [67.48%]
       561,945,984 instructions              #    1.18  insns per cycle        
                                             #    0.36  stalled cycles per insn [85.07%]
       132,785,648 branches                  #  701.913 M/sec                   [84.93%]
         1,847,286 branch-misses             #    1.39% of all branches         [84.62%]

       0.169685736 seconds time elapsed

$ perf stat ./dxexmachina_devstone opdevs 2 2 3
 Performance counter stats for './dxexmachina_devstone opdevs 2 2 3':

        258.435415 task-clock (msec)         #    0.924 CPUs utilized          
             1,201 context-switches          #    0.005 M/sec                  
                26 cpu-migrations            #    0.101 K/sec                  
             3,836 page-faults               #    0.015 M/sec                  
       613,419,141 cycles                    #    2.374 GHz                     [83.93%]
       316,632,123 stalled-cycles-frontend   #   51.62% frontend cycles idle    [85.01%]
       139,998,026 stalled-cycles-backend    #   22.82% backend  cycles idle    [67.52%]
       615,911,601 instructions              #    1.00  insns per cycle        
                                             #    0.51  stalled cycles per insn [84.95%]
       144,090,841 branches                  #  557.551 M/sec                   [85.11%]
         2,820,791 branch-misses             #    1.96% of all branches         [78.96%]

       0.279720722 seconds time elapsed

$ perf stat ./dxexmachina_devstone cpdevs 2 2 3
 Performance counter stats for './dxexmachina_devstone cpdevs 2 2 3':

        257.663639 task-clock (msec)         #    0.851 CPUs utilized          
               911 context-switches          #    0.004 M/sec                  
                19 cpu-migrations            #    0.074 K/sec                  
             3,837 page-faults               #    0.015 M/sec                  
       663,670,571 cycles                    #    2.576 GHz                     [82.64%]
       351,293,037 stalled-cycles-frontend   #   52.93% frontend cycles idle    [82.43%]
       197,270,653 stalled-cycles-backend    #   29.72% backend  cycles idle    [65.25%]
       643,477,063 instructions              #    0.97  insns per cycle        
                                             #    0.55  stalled cycles per insn [84.74%]
       150,533,837 branches                  #  584.226 M/sec                   [87.77%]
         2,968,904 branch-misses             #    1.97% of all branches         [82.67%]

       0.302693494 seconds time elapsed
\end{Verbatim}
\paragraph{\textsc{PythonPDEVS}}
\begin{Verbatim}[fontsize=\small]
$ perf stat ./dxexmachina_devstone classic 2 3
 Performance counter stats for 'python ./DEVStone/experiment.py default 2 false':

        670.501187 task-clock (msec)         #    0.959 CPUs utilized          
               169 context-switches          #    0.252 K/sec                  
                22 cpu-migrations            #    0.033 K/sec                  
             4,962 page-faults               #    0.007 M/sec                  
     1,828,074,336 cycles                    #    2.726 GHz                     [82.96%]
       983,729,421 stalled-cycles-frontend   #   53.81% frontend cycles idle    [83.51%]
       600,902,567 stalled-cycles-backend    #   32.87% backend  cycles idle    [67.59%]
     1,691,525,594 instructions              #    0.93  insns per cycle        
                                             #    0.58  stalled cycles per insn [84.05%]
       380,934,259 branches                  #  568.134 M/sec                   [83.42%]
        15,797,482 branch-misses             #    4.15% of all branches         [82.57%]

       0.699383675 seconds time elapsed

$ perf stat ./dxexmachina_devstone opdevs 2 2 3
 Performance counter stats for 'mpirun -np 2 python ./DEVStone/experiment.py default 2 false':

        692.210888 task-clock (msec)         #    0.330 CPUs utilized          
               909 context-switches          #    0.001 M/sec                  
                99 cpu-migrations            #    0.143 K/sec                  
             8,728 page-faults               #    0.013 M/sec                  
     1,801,930,537 cycles                    #    2.603 GHz                     [82.89%]
       929,771,707 stalled-cycles-frontend   #   51.60% frontend cycles idle    [82.88%]
       581,151,775 stalled-cycles-backend    #   32.25% backend  cycles idle    [68.37%]
     1,800,262,136 instructions              #    1.00  insns per cycle        
                                             #    0.52  stalled cycles per insn [84.52%]
       403,185,223 branches                  #  582.460 M/sec                   [84.04%]
        13,159,671 branch-misses             #    3.26% of all branches         [82.92%]

       2.099629915 seconds time elapsed
\end{Verbatim}
\subsection{PHOLD}
\paragraph{\textsc{DEVS Ex Machina}}
\begin{Verbatim}[fontsize=\small]
$ perf stat ./dxexmachina_phold classic 1 10 0 10
 Performance counter stats for './dxexmachina_phold classic 1 10 0 10':

         12.815382 task-clock (msec)         #    1.081 CPUs utilized          
                11 context-switches          #    0.858 K/sec                  
                25 cpu-migrations            #    0.002 M/sec                  
               712 page-faults               #    0.056 M/sec                  
        27,657,920 cycles                    #    2.158 GHz                     [64.22%]
        13,070,687 stalled-cycles-frontend   #   47.26% frontend cycles idle    [75.41%]
         8,563,853 stalled-cycles-backend    #   30.96% backend  cycles idle   
        31,010,010 instructions              #    1.12  insns per cycle        
                                             #    0.42  stalled cycles per insn
         6,957,728 branches                  #  542.920 M/sec                  
           122,049 branch-misses             #    1.75% of all branches         [69.03%]

       0.011857863 seconds time elapsed

$ perf stat ./dxexmachina_phold opdevs 2 1 10 0 10
 Performance counter stats for './dxexmachina_phold opdevs 2 1 10 0 10':

         58.345592 task-clock (msec)         #    0.103 CPUs utilized          
               235 context-switches          #    0.004 M/sec                  
                24 cpu-migrations            #    0.411 K/sec                  
               844 page-faults               #    0.014 M/sec                  
        86,073,977 cycles                    #    1.475 GHz                     [79.15%]
        41,603,661 stalled-cycles-frontend   #   48.33% frontend cycles idle    [87.74%]
        23,030,593 stalled-cycles-backend    #   26.76% backend  cycles idle    [54.61%]
        93,532,407 instructions              #    1.09  insns per cycle        
                                             #    0.44  stalled cycles per insn [79.51%]
        20,525,076 branches                  #  351.785 M/sec                   [88.38%]
           426,419 branch-misses             #    2.08% of all branches         [93.36%]

       0.564278040 seconds time elapsed

$ perf stat ./dxexmachina_phold cpdevs 2 1 10 0 10
 Performance counter stats for './dxexmachina_phold cpdevs 2 1 10 0 10':

         45.097489 task-clock (msec)         #    0.453 CPUs utilized          
                46 context-switches          #    0.001 M/sec                  
                13 cpu-migrations            #    0.288 K/sec                  
               856 page-faults               #    0.019 M/sec                  
        68,793,923 cycles                    #    1.525 GHz                     [67.00%]
        40,973,357 stalled-cycles-frontend   #   59.56% frontend cycles idle    [82.61%]
        15,619,202 stalled-cycles-backend    #   22.70% backend  cycles idle    [71.88%]
        56,478,836 instructions              #    0.82  insns per cycle        
                                             #    0.73  stalled cycles per insn [91.10%]
        13,722,715 branches                  #  304.290 M/sec                   [91.32%]
           279,636 branch-misses             #    2.04% of all branches         [90.09%]

       0.099622680 seconds time elapsed
\end{Verbatim}
\paragraph{\textsc{PythonPDEVS}}
\begin{Verbatim}[fontsize=\small]
$ perf stat ./dxexmachina_phold classic 1 10 0 10
 Performance counter stats for 'python ./benchmarks/experiment.py':

        671.561546 task-clock (msec)         #    0.998 CPUs utilized          
                97 context-switches          #    0.144 K/sec                  
                18 cpu-migrations            #    0.027 K/sec                  
             2,974 page-faults               #    0.004 M/sec                  
     1,916,305,122 cycles                    #    2.854 GHz                     [83.35%]
     1,005,734,716 stalled-cycles-frontend   #   52.48% frontend cycles idle    [83.34%]
       499,103,479 stalled-cycles-backend    #   26.05% backend  cycles idle    [66.69%]
     2,060,775,055 instructions              #    1.08  insns per cycle        
                                             #    0.49  stalled cycles per insn [83.34%]
       345,264,964 branches                  #  514.123 M/sec                   [83.89%]
        12,293,319 branch-misses             #    3.56% of all branches         [82.87%]

       0.673223302 seconds time elapsed

$ perf stat ./dxexmachina_phold opdevs 2 1 10 0 10
 Performance counter stats for 'mpirun -np 2 python ./benchmarks/experiment.py':

        866.680461 task-clock (msec)         #    0.800 CPUs utilized          
               929 context-switches          #    0.001 M/sec                  
                83 cpu-migrations            #    0.096 K/sec                  
             7,466 page-faults               #    0.009 M/sec                  
     2,326,077,914 cycles                    #    2.684 GHz                     [83.64%]
     1,283,069,850 stalled-cycles-frontend   #   55.16% frontend cycles idle    [83.95%]
       940,158,530 stalled-cycles-backend    #   40.42% backend  cycles idle    [65.55%]
     2,325,361,346 instructions              #    1.00  insns per cycle        
                                             #    0.55  stalled cycles per insn [82.69%]
       402,983,176 branches                  #  464.973 M/sec                   [83.67%]
        15,476,432 branch-misses             #    3.84% of all branches         [84.05%]

       1.083881333 seconds time elapsed
\end{Verbatim}
